Search for Spatially Extended \fermi-LAT Sources Using Two Years of Flight Data

### 2011-09-15 extended_catalog/paper/extended_catalog.tex
update draft
udpate code

@@

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\km}{\text{km}\xspace}
\newcommand{\phflux}{\ensuremath{\text{ph}\ \text{cm}^{-2}\,\text{s}^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{inc}}}\xspace}

@@

@@

\subsection{Dual Localization}
\label{dual_localization_method}

There is a degeneracy between a spatially extended source and multiple
point-like  sources separated by angular distances comparable to or smaller than
the size of the LAT PSF.  To assess the possibility of source confusion,
we use \pointlike to simultaneously fit the position of two point-like sources
in the region of extended source candidates.

We define \tsinc as twice the increase in the log of the likelihood fitting the
region as two point-like sources compared to fitting the region as one point-like
source: 
\begin{equation}
  \tsinc=2\log(\likelihood_\text{2pts}/\likelihood_\text{ps}).
\end{equation} 
\tsinc can not be directly compared to \tsext to see which
model is more significant because the models are not nested
\citep{statistics_with_care}. Even though the comparison of \tsext with
\tsinc is not a calibrated test, we find the cases $\tsinc \ll \tsext$
and $\tsinc\gg\tsext$ suggestive and we only consider a source to be
extended if $\tsext>\tsinc$.
Similar to the case of extended sources described in Section~\ref{gtlike_crosscheck},
the spectra of the two point-like sources can be refit using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions found using \pointlike.

@@

We test the false detection probability of for the
extended source test by fitting the extension of
point-like sources.
\cite{mattox_egret} discuss that the test statistic distribution
for a likelihood ratio test on the existence of a source at
a given position is 
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\onehalf(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of Equation \ref{ts_ext_distribution} is
due to the null hypothesis (source flux $\Phi=0$) residing
on the edge of parameter space and the model hypothesis
adding a single degree of freedom
and leads to the often quoted result that the square root
of the test statistic is the number of $\sigma$ of
the detection. It is plausible to
expect a similar distribution of the test statistic
in the test for source extension since the same conditions
apply (with the source flux $\Phi$ replaced by the source radius $r$ and
$r<0$ being unphysical).
We performed Monte Carlo
simulations
to calculate empirical distributions for $\tsext$ and
we compared them to Equation~\ref{ts_ext_distribution}.

@@

@@

We estimate a second systematic error on the extension of a source
due to uncertainty in out model of the Galactic diffuse emission by
using an alternative diffuse model based
upon GALPROP\footnote{GALPROP is a software package for calculating the
Galactic $\gamma$-ray emission based on a model of cosmic-ray propagation
in the Galaxy. See \url{http://galprop.stanford.edu/} for details
and references.} and used in the LAT analysis of the isotropic diffuse
emission \citep{isotropic_lat}.  The intensities of various components
of the Galactic diffuse emission were fitted individually using a
spatial distribution of the intensities within the ROI as predicted by
the model.  We distinguished contributions from CR interactions with the
molecular hydrogen, the atomic+ionized hydrogen, residual gas traced
by dust \citep{isabelle_dark_gass}, and the interstellar radiation
field. We further split the contributions from interactions with molecular
and atomic hydrogen to the Galactic diffuse emission according to the
distance from the Galactic center in which they are produced. Hence, we
replaced the standard diffuse emission model by 18 individually fitted
templates to describe individual components of the diffuse emission.
A similar crosscheck was used in the LAT Collaboration's analysis of 
RX\,J1713.7$-$3946 \citep{rx_j1713_lat}.

@@

Of the 885 clean AGN, we selected the 783 of these 2FGL sources which
were significantly-detected above 1 \gev and fit each of them for extension.
The cumulative density of \tsext for these AGN compared to the
$\chi^2/2$ distribution of Equation~\ref{ts_ext_distribution}
is
shown in Figure~\ref{agn_ts_ext}. 
The \tsext distribution for AGN shows good agreement with the
theoretical distribution, but two sources had $\tsext>10$.  One was due
to the incorrect removal of a nearby catalog sources from
the sky model (see
Section~\ref{extended_source_search_method}) and the other was due to a
failure of convergence of the point-like source hypothesis.  This result demonstrates
that we can use \tsext as a measure of the statistical significance of
the detection of the extension of a source.

@@

This extended sources is coincident with 2CG\,078+01 and
3EG\,J2020+4017 \citep{second_cos_b_catalog,third_egret_catalog}
and also with the $\gamma$-Cygni SNR. This $\gamma$-ray source has
been speculated to originate from the interaction of accelerated
particles in the SNR interacting with dense molecular clouds
\citep{pollock_1985,gaisser_1998},  but this association was disfavored
when the LAT \gev emission from this source was detected to be pulsed
(PSR\,J2021+4026, \cite{first_lat_pulsar_cat}).

@@

@@

\begin{figure}
    \ifcolorfigure
    \plotone{mc_plots/index_sensitivity_color.eps}
    \else
    \plotone{mc_plots/index_sensitivity_bw.eps}
    \fi
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/monte_carlo/sensitivity/v12/plot/plot_vs_index.py
    \caption{
    The detection threshold to resolve a
    uniform surface brightness extended source for a two-year exposure.  All sources
    have an assumed power-law spectrum and the different colors correspond
    to different simulated spectral indices.  The solid line is the
    detection threshold using photons with energies between 100 \mev and 100
    \gev while the dashed line is the threshold using 
     photons with energies between 1 \gev and 100 \gev.
    }\label{index_sensitivity}
  \end{figure}

@@

@@

\clearpage
\begin{figure}
    \ifcolorfigure
      \plotone{source_plots/agn_color.eps}
    \else
      \plotone{source_plots/agn_bw.eps}
    \fi
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/agn/v6/agn.py
    \caption{The cumulative density of \tsext for the 783 clean
    AGN in 2LAC that were significant above 1 \gev calculated with \pointlike (dashed line
    colored blue in the electronic version)
    and with \gtlike (solid line colored red).  AGN are too far
    and too small to be resolved by the LAT. Therefore, the cumulative
    density of $\tsext$ is expected to follow a $\chi^2/2$ distribution
    (Equation~\ref{ts_ext_distribution}, the dash-dotted line colored
    red).
    }\label{agn_ts_ext}
  \end{figure}

>>>>

% Changes from v2.0 to v3.0

% * Move section 'Dual Localization' to its own
%   section after 'Extended Source Detection Threshold'
%   and rename it 'Testing Against Source Confusion'
%   -> Significantly expanded the section with a new
%      monte carlo study fitting extended sources as
%      two point sources, and vice versa
%
% * replace all occurances of ts_inc with ts_2pts
%%%%%%%%
% * Added acknowlegements section to end of the paper
% * Move HESS SEDs to earlier in paper to fix order.
% * Fix some occurances of chi^2 to correctly be chi^2_1
% * Better description of how the alternate diffuse model
%   is different from galprop
% * Add galprop citation
% * Add AGILE Gamma Cygni Pulsar citation.

@@

% some units
\newcommand{\kev}{\text{Kev}\xspace}
\newcommand{\mev}{\text{MeV}\xspace}
\newcommand{\gev}{\text{GeV}\xspace}
\newcommand{\tev}{\text{TeV}\xspace}
\newcommand{\s}{\text{s}\xspace}
\newcommand{\ph}{\text{ph}\xspace}
\newcommand{\cm}{\text{cm}\xspace}
\newcommand{\km}{\text{km}\xspace}
\newcommand{\phflux}{\ensuremath{\text{ph}\,\text{cm}^{-2}\,\text{s}^{-1}}\xspace}
\newcommand{\tsext}{{\ensuremath{\text{TS}_{\text{ext}}}}\xspace}
\newcommand{\tsinc}{\ensuremath{\text{TS}_{\text{2pts}}}\xspace}

@@

%\newcommand{\newtext}[1]{{\textbf #1}}
\newcommand{\newtext}[1]{{\bfseries \color{red}#1}}

@@

@@

We tested the false detection probability for the extended source test
by fitting the extension of point-like sources.  \cite{mattox_egret}
discuss that the test statistic distribution for a likelihood ratio test
on the existence of a source at a given position is
\begin{equation}\label{ts_ext_distribution}
  P(\ts)=\onehalf(\chi^2_1(\ts)+\delta(\ts)).
\end{equation}
The particular form of Equation \ref{ts_ext_distribution} is due to the
null hypothesis (source flux $\Phi=0$) residing on the edge of parameter
space and the model hypothesis adding a single degree of freedom and
leads to the often quoted result that the square root of the test
statistic is the number of $\sigma$ of the detection. It is plausible
to expect a similar distribution of the test statistic in the test for
source extension since the same conditions apply (with the source flux
$\Phi$ replaced by the source radius $r$ and $r<0$ being unphysical).
We performed Monte Carlo simulations to calculate empirical distributions
for $\tsext$ and we compared them to Equation~\ref{ts_ext_distribution}.

@@

\section{Testing Against Source Confusion}
\label{dual_localization_method}

\newtext{
It is impossible to discriminate with LAT data exclusively between a
spatially extended source and multiple point-like sources separated by
angular distances comparable to or smaller than the size of the LAT PSF.
To assess the plausibility of source confusion, we used \pointlike to
simultaneously fit the position and spectrum of two point-like sources
in the region of extended source candidates.

We define \tsinc as twice the increase in the log of the likelihood
fitting the region as two point-like sources compared to fitting the
region as one point-like source:
\begin{equation}
  \tsinc=2\log(\likelihood_\text{2pts}/\likelihood_\text{ps}).
\end{equation} 
\tsinc cannot be quantitatively compared to \tsext using a
simple likelihood ratio test to evaluate which model provides the
most significant improvement because the models are not nested
\citep{statistics_with_care}.  Even though the comparison of \tsext
with \tsinc is not a calibrated test, $\tsext>\tsinc$ indicates that
the likelihood for the extended source hypothesis is higher than for
two point-like sources and we only consider a source to be extended
if $\tsext>\tsinc$.

We assessed the power of the test comparing \tsinc to \tsext to correctly
distinguish between source confusion and a spatially extended source using
a Monte Carlo study.  We simulated two point sources and fit them as both
an extended source and two point-like sources.  We then simulated two
point sources and fit them with the same hypothesis. We could therefore
compare \tsinc and \tsext for the case of there being actually two point
sources and of there being extended source.  All sources were simulated
using the same two-year time range included in 2FGL and all fit values
were calculated using \pointlike.

We did this study first in the energy range from 1 \gev to 100 \gev by
simulating extended sources of flux $4\times10^{-9} \phflux$ in the 1
\gev to 100 \gev energy range and with a power-law spectral model with
spectral index 2.  This flux was picked to be representative of the
new extended sourced that were discovered by our subsequent analysis
looking at energies between 1 \gev and 100 \gev
(see section~\ref{new_ext_srcs_section}).
We simulated these sources using a uniform surface brightness spatial
model with extensions varying
up to $1\degree$.  Figure~\ref{confusion_plot}a shows 
the distribution of \tsinc and \tsext as a
function of the simulated extension of the source
for 200 statistically independent simulations.

Figure~\ref{confusion_plot}c shows the same analysis but when fitting
simulations of two point-like sources each with power-law spectral models
with a spectral index of 2.  Here, the two point-like sources are each given half
of the flux of the extended source ($2\times10^{-9} \phflux$ integrated
from 1 \gev to 100 \gev) to ensure that the two point-like sources predict
the same number of photons but with a different spatial shape.
Finally, figure~\ref{confusion_plot}e shows the same analysis but with
one point source given a spectral index of 1.5 and the other point source
given an index of 2.5, but with each point source still have an integral
1 \gev to 100 \gev flux of $2\times10^{-9} \phflux$.

The same three plots are shown in 
figure~\ref{confusion_plot}b, figure~\ref{confusion_plot}d,
and figure~\ref{confusion_plot}f but this time when analyzing
a source of flux $10^{-9} \phflux$ (integrated from 10 \gev to 100 \gev) 
analyzed only in the 10 \gev to 100 \gev energy range.
This spectrum is typical 
of the new extended sources discovered using only photons with energies
between 1 \gev and 100 \gev (see section~\ref{new_ext_srcs_section}).

Several interesting conclusions can be made from this Monte
Carlo study. The overall behavior of these plots is what one would
expect. \tsext is on average greater than \tsinc when fitting simulated
extended sources whereas \tsinc is greater than \tsext when fitting two
simulated point sources.  Furthermore, the distributions of \tsext and
\tsinc separate better when the extended source is bigger and when the
two point sources are farther apart.

On the other hand, the plots show that for extended sources with typical
sizes $\sim0\fdg5$, one can often obtain almost as good a likelihood
increase when fitting a simulated extended source as two point sources and
that \tsinc is often not much larger.  This makes sense because although
two point-like sources represent an incorrect spatial model when fitting
an extended sources, the two point-like sources have three more degrees
of freedom than the one extended source and can therefore easy model
statistical fluctuations in the counts distribution to create almost as
good a fit.  This effect is most pronounced when using energies from 1
\gev to 100 \gev where the PSF is even larger.  From this, we conclude
that if we have an extended source candidate where \tsinc is smaller
than but comparable to \tsext, this is not good evidence that we are
most likely seeing instead source confusion.

When testing the two simulated point-like sources for extension, our
test did a better job at separating the distribution of \tsinc and
\tsext, especially for larger separation distances.  This says that
when there are two point-like sources instead of one extended source,
we really are less likely to confuse the emission as looking more like
a spatially extended source.

Talk about how comparing the spectral index is often a
difficult way to distinguish between two point sources
and one extended source. Discuss figure \ref{index_difference}.

From study, we see that \dots

Conclusions from the MC study:
* TSinc can be very large even for a true extended source.

This shows that the LAT can not uniquiely determinte what featuers
we see in the data. The 2FGL catalog's method of fitting all residuals
as new point sources is incorrect as is assuming everything is an
extended source an not source confusion. A more nuanced, but somewhat arbitrary,
criterion must be used on a case by case basis using multiwavelenth
information to decide whether we believe our best fit model is 
composed of multiple separate sources or one spatially extended source.

Even so, if the LAT detects a source as being extended, it could
likely in the future be resolved into multiple point sources.
This is actually almost inevitable because of the large density of
sources in the galactic plane which could all be \gev emitters.

Even though this dedicated monte carlo study was performed
using \pointlike, for the analysis of LAT data we took the best
fit positions of the two point-like sources and refit them using \gtlike.
We quote the spectral values obtained from \gtlike using the best fit
positions of the two point-like sources found using \pointlike.

@@

\newtext{
We estimated a second systematic error on the extension of a source
due to uncertainty in our model of the Galactic diffuse emission by
using an alternative model which takes as input templates
calculated by
GALPROP\footnote{GALPROP is a software package for calculating the
Galactic $\gamma$-ray emission based on a model of cosmic-ray propagation
in the Galaxy \citep{galprop1998,galprop2011}. 
See also \url{http://galprop.stanford.edu/} for details.} 
but then fits each template locally in
the surrounding region.
The particular GALPROP model that was used as input is described in
the LAT analysis of the isotropic diffuse
emission \citep{isotropic_lat}.  
}
The intensities of various components
of the Galactic diffuse emission were fitted individually using a
spatial distribution of the intensities within the ROI as predicted by
the model.  We distinguished contributions from CR interactions with the
molecular hydrogen, the atomic+ionized hydrogen, residual gas traced
by dust \citep{isabelle_dark_gass}, and the interstellar radiation
field. We further split the contributions from interactions with molecular
and atomic hydrogen to the Galactic diffuse emission according to the
distance from the Galactic center in which they are produced. Hence, we
replaced the standard diffuse emission model by 18 individually fitted
templates to describe individual components of the diffuse emission.
A similar crosscheck was used in the LAT Collaboration's analysis of 
RX\,J1713.7$-$3946 \citep{rx_j1713_lat}.

@@

Of the 885 clean AGN, we selected the 783 of these 2FGL sources which
were significantly-detected above 1 \gev and fit each of them for extension.
The cumulative density of \tsext for these AGN compared to the
$\chi^2_1/2$ distribution of Equation~\ref{ts_ext_distribution}
is
shown in Figure~\ref{agn_ts_ext}. 
The \tsext distribution for AGN shows good agreement with the
theoretical distribution, but two sources had $\tsext>10$.  One was due
to the incorrect removal of a nearby catalog sources from
the sky model (see
Section~\ref{extended_source_search_method}) and the other was due to a
failure of convergence of the point-like source hypothesis.  This result demonstrates
that we can use \tsext as a measure of the statistical significance of
the detection of the extension of a source.

@@

This extended sources is coincident with 2CG\,078+01 and
3EG\,J2020+4017 \citep{second_cos_b_catalog,third_egret_catalog}
and also with the $\gamma$-Cygni SNR. This $\gamma$-ray source has
been speculated to originate from the interaction of accelerated
particles in the SNR interacting with dense molecular clouds
\citep{pollock_1985,gaisser_1998},  but this association was disfavored
when the LAT \gev emission from this source was detected to be pulsed
(PSR\,J2021+4026, \cite{first_lat_pulsar_cat}). 
\newtext{
This
pulsar was also observed by AGILE \citep{gamma_cygni_agile}.
}

@@

\newtext{

The \textit{Fermi} LAT Collaboration acknowledges generous ongoing support
from a number of agencies and institutes that have supported both the
development and the operation of the LAT as well as scientific data analysis.
These include the National Aeronautics and Space Administration and the
Department of Energy in the United States, the Commissariat \`a l'Energie Atomique
and the Centre National de la Recherche Scientifique / Institut National de Physique
Nucl\'eaire et de Physique des Particules in France, the Agenzia Spaziale Italiana
and the Istituto Nazionale di Fisica Nucleare in Italy, the Ministry of Education,
Culture, Sports, Science and Technology (MEXT), High Energy Accelerator Research
Organization (KEK) and Japan Aerospace Exploration Agency (JAXA) in Japan, and
the K.~A.~Wallenberg Foundation, the Swedish Research Council and the
Swedish National Space Board in Sweden.

Additional support for science analysis during the operations phase is gratefully
acknowledged from the Istituto Nazionale di Astrofisica in Italy and the Centre National d'\'Etudes Spatiales in France.

This research has made use of
pywcsgrid2, an open-source plotting package for
Python\footnote{http://leejjoon.github.com/pywcsgrid2/}. The authors acknowledge the
use of HEALPix\footnote{http://healpix.jpl.nasa.gov/} \citep{healpix}.

@@

\clearpage
\begin{figure}
    \ifcolorfigure
    \plotone{mc_plots/index_sensitivity_color.eps}
    \else
    \plotone{mc_plots/index_sensitivity_bw.eps}
    \fi
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/monte_carlo/sensitivity/v12/plot/plot_vs_index.py
    \caption{
    The detection threshold to resolve a
    uniform surface brightness extended source for a two-year exposure.  All sources
    have an assumed power-law spectrum and the different colors correspond
    to different simulated spectral indices.  The solid line is the
    detection threshold using photons with energies between 100 \mev and 100
    \gev while the dashed line is the threshold using 
     photons with energies between 1 \gev and 100 \gev.
    }\label{index_sensitivity}
  \end{figure}

@@

\clearpage
\begin{figure}
    \ifcolorfigure
    \plotone{mc_plots/confusion_plot_color.eps}
    \else
    \plotone{mc_plots/confusion_plot_bw.eps}
    \fi
    % this plot came from 
    \caption{Testing Confusion.
    }\label{confusion_plot}
  \end{figure}

\clearpage
\begin{figure}
    \ifcolorfigure
    \plotone{mc_plots/index_difference_color.eps}
    \else
    \plotone{mc_plots/index_difference_bw.eps}
    \fi
    % this plot came from 
    \caption{Index difference.
    }\label{index_difference}
  \end{figure}

@@

\clearpage
\begin{figure}
    \ifcolorfigure
      \plotone{source_plots/agn_color.eps}
    \else
      \plotone{source_plots/agn_bw.eps}
    \fi
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/agn/v6/agn.py
    \caption{The cumulative density of \tsext for the 783 clean
    AGN in 2LAC that were significant above 1 \gev calculated with \pointlike (dashed line
    colored blue in the electronic version)
    and with \gtlike (solid line colored red).  AGN are too far
    and too small to be resolved by the LAT. Therefore, the cumulative
    density of $\tsext$ is expected to follow a $\chi^2_1/2$ distribution
    (Equation~\ref{ts_ext_distribution}, the dash-dotted line colored
    red).
    }\label{agn_ts_ext}
  \end{figure}
